Sample Length at Which Block Bootstrapping is Effective for Estimation of Parameter of Time Series

### 2023-02-16 Manuscript/manuscript.tex
Line Width

While the aforementioned studies have looked into the necessary sample size for various bootstrap methods, the goal of this study is to find a threshold or range of sample sizes at which block bootstrap 
is an effective method for estimating a parameter specifically in the context of a time series. The method can be considered effective if approximately 1 - $\alpha$/2 \% of 1 - $\alpha$/2 \% of confidence intervals created from samples of the same size recover the parameter. This study will focus on estimating parameters of an AR(1) process with two types of confidence intervals from the literature, \citep{diciccio1996bootstrap}  \citep{rice2006mathematical} which will be reviewed in Section 2.

%At the base level, we are block-bootstrap an auto-regressive process with true mean 0, and some AR(1) coefficient.

%Bootstrap is the procedure of creating new pseudo-samples of the same size by resampling observations with replacement from the original sample. Basic bootstrap works well for samples
% that are do not have a dependence factor. However, for a time series such as an auto-regressive or moving-average process, a different procedure is required to account for the time dependence. In such a situation, instead of resampling observations, blocks of the original sample can be resampled (these blocks must be of a certain length by which the sample length is divisible) to create a pseudo-sample. Point estimates of a parameter calculated from many block bootstrapped pseudo-samples are then used to create a distribution to create an interval estimation of the parameter.

%In this experiment, the mean, auto-correlation function (with a maximum lag of 1), and variance of a 1000 block-bootstrapped time series samples are computed, and 95\% confidence intervals of these parameters is created.

%The various ways which these intervals can be constructed are described in \citet{diciccio1996bootstrap} and \citet{rice2006mathematical}. They will be discussed later in a review of the block bootstrap method. For the purposes of this study (in which the true parameters are known), in order to establish whether or not the method is effective, the block bootstrap of a simulated sample can be replicated many times, and the proportion of confidence intervals that recover the true known parameter (coverage rate) is recorded. The goal of this is simply to observe how successful the block bootstrap procedure is at estimating a parameter. The key variable being observed was, the sample length (n). It is known that as n increases, block bootstrap should become a more accurate and precise method for estimation(the coverage rates will approach the theoretical confidence level and the interval widths will decrease). The question is what is the minimum sample length necessary to reach a coverage rate of 1 - $\alpha$/2 \%.

% In this experiment, there are certain variables that are expected to affect the performance of block bootstrap. As the the time dependence (AR or MA coefficient) of the time series increases, we expect the coverage rates of the confidence intervals to decrease.The coverage rates are also affected by the size of the blocks - more specifically,how the size of the blocks (l) relates to the size of the time series (n) - and whether or not they overlap. It is known that as the size of the time series increases, the optimal block length should increase, and the ratio of the block length to the time series length should decrease. l = n$^{1/3}$ is typically considered the optimal block length according to asymptotic theory. \citep{buhlmann1999block} Therefore, simulations in this study were conducted using this optimal function of l = n$^{1/3}$.

%In order to solve the question of what sample length is necessary for effective block bootstrap while still accounting for all these factors, the simulations were repeated for different variations of the MA coefficient, target parameter, and non-moving vs moving method. For each of these variations, results for block bootstrap estimation of mean, MA coefficient, and standard deviation were recorded for a wide range of sample length.

@@

In regular bootstrap procedure, the new pseudo-sample of the same size as the original sample would be created by simply resampling observations from the original sample with replacement. However, in the case of a time series, in order to account for the temporal dependence, the time series can be split into blocks, typically of the same size. The block should be of size $l$ large enough for each pseudo-sample to exhibit some temporal dependence, yet small enough for there to be some variability between each pseudo-sample. Ideally, as $n$ increases, $l$ should also increase, but $l / n$ should decrease. To achieve this, $l$ is often assigned a value as a function of $n$. A common function that is considered the best by much previous literature is $l = \lceil n^{1/3} \rceil$ \citep{buhlmann1999block}. The designer of the study can choose whether the blocks overlap or not. If the blocks do not overlap, it is called a non-moving block bootstrap. In this case, the original sample is split evenly into blocks, which will then be resampled to create new pseudo-samples. If the blocks do overlap, it is called a moving block bootstrap. Blocks are resampled randomly from the original sample, but they do not have to be from a set of evenly spaced blocks. For moving block bootstrap, if $l \vert n$, a bootstrapped pseudo-sample is created by taking a sample of $n / l$ blocks of size $l$ with replacement to create a new pseudo-sample of size $n$. If $l does not divide n$, the pseudo-sample is created by taking a sample of $\lfloor n / l \rfloor$ blocks of size $l$ and one block of size $(n \;\mathrm{mod})\; l$. An estimate $\hat\theta_n^*$ is computed from the pseudo-sample. This procedure can be repeated many times to create a distribution of $\hat\theta_n^*$.

@@

Thus, there are two types of intervals used for estimation in this study: $CI_{\hat{\theta}_{n}}$ and $CI_{BCA, \hat{\theta}_{n}}$. Note that both of these intervals are centered around $\hat{\theta}_{n}$ and asymmetric.

@@

Each of these 6 intervals either recovers $\theta$ or does not. Remember that 95\% of many of a certain type of 95\% confidence intervals constructed from samples of the same $n$ should recover $\theta$. To see if this is generally or approximately the case, one can run 1000 replications of the block bootstrap procedure, and for each interval-$\theta$ pair, the rate at which that type of interval recovers $\theta$ can be recorded.

If $\theta$ is not covered at that $n$, a larger sample size must be necessary. $n = $100, 200, 300, 400, and 500, 600, 700 were used in this study.

@@

Only $CI_{\hat{\theta}_{n}}$ managed to solve the deterioration problem with respect to recovering $\phi$, indicating that this interval should be used as opposed to $CI_{BCA, \hat{\theta}_{n}}$ when estimating $\phi$.

@@

The motivation for the study is the idea that block bootstrap procedure will perfectly estimate a parameter of a time series given an infinitely large sample. The goal for this study was to find what is a good enough finite sample length for block bootstrap estimation to recover the parameter of a time series at an acceptable rate. Assuming that $CI_{\hat{\theta}_{n}}$ is used, The results of this study suggest that even for estimating $\mu$, an $n$ of around 500 may be necessary if $\phi$ is unknown. An $n$ of around 400 may be necessary to estimate $\sigma$ is $\phi$ is unknown. If $\phi$ is already known, a lesser $n$ may be adequate to estimate these parameters. Lastly, to estimate $\phi$, an $n$ of around 400 may be necessary. This information could prove to be useful for researching using block bootstrap estimation of time series in domains such as econometrics. Future studies could investigate the sample size needed to estimate or test an hypothesis about other forms of serially dependent data such as an MA process.

>>>>

While the aforementioned studies have looked into the necessary sample size for various bootstrap methods, 
the goal of this study is to find a threshold or range of sample sizes at which block bootstrap 
is an effective method for estimating a parameter specifically in the context of a time series. 
The method can be considered effective if approximately 1 - $\alpha$/2 \% of 1 - $\alpha$/2 \% of confidence intervals 
created from samples of the same size recover the parameter. This study will focus on estimating parameters of an 
AR(1) process with two types of block bootstrap confidence intervals from the literature, \citep{diciccio1996bootstrap}  \citep{rice2006mathematical} 
which is reviewed in Section 2. Section 3 offers an explanation of the simulation design and results. 
Section 4 discusses the results and how these findings could be useful.

@@

In basic bootstrap procedure, the new pseudo-sample of size $n$ would be created by simply resampling observations from 
the original sample with replacement. However, in the case of a time series, in order to account for the temporal dependence, 
the time series can be split into blocks, typically of the same size. The block should be of size $l$ large enough for each pseudo-sample
 to exhibit some temporal dependence, yet small enough for there to be some variability between each pseudo-sample. 
 Ideally, as $n$ increases, $l$ should also increase, but $l / n$ should decrease. To achieve this, $l$ is often assigned 
 a value as a function of $n$. A common function that is considered the best by much previous literature is $l = \lceil n^{1/3} \rceil$ \citep{buhlmann1999block}. 
 The designer of the study can choose whether the blocks overlap or not. If the blocks do not overlap, it is called a non-moving block bootstrap. 
 In this case, the original sample is split evenly into blocks, which will then be resampled to create new pseudo-samples. 
 If the blocks do overlap, it is called a moving block bootstrap. Blocks are resampled randomly from the original sample, 
 but they do not have to be from a set of evenly spaced blocks. For moving block bootstrap, if $l \vert n$, a bootstrapped 
 pseudo-sample is created by taking a sample of $n / l$ blocks of size $l$ with replacement to create a new pseudo-sample of size $n$. 
 If $l does not divide n$, the pseudo-sample is created by taking a sample of $\lfloor n / l \rfloor$ blocks of size $l$ 
 and one block of size $(n \;\mathrm{mod})\; l$. An estimate $\hat\theta_n^*$ is computed from the pseudo-sample. 
 This procedure can be repeated many times to create a distribution of $\hat\theta_n^*$.

@@

Thus, there are two types of intervals that can be constructed for block bootstrap estimation in this study: 
$CI_{\hat{\theta}_{n}}$ and $CI_{BCA, \hat{\theta}_{n}}$. Note that both of these intervals are centered around $\hat{\theta}_{n}$ and asymmetric.

@@

Each of these 6 intervals either recovers $\theta$ or does not. Remember that 95\% of many of a certain type of 95\% confidence intervals 
constructed from samples of the same $n$ should recover $\theta$. To see if this is generally or approximately the case, one can run 
1000 replications of the block bootstrap procedure, and for each interval-$\theta$ pair, the rate at which that type of interval recovers 
$\theta$ can be recorded. The coverage rate is, of course, a point estimate of a proportion, so a 95\% confidence interval of this coverage proportion 
(with $n_{CI} = 1000$) can be constructed, and if the proportion .95 is included in the interval, the block bootstrap method is likely working well.

If $\theta$ is not covered at that $n$, a larger sample size must be necessary. $n = $100, 200, 300, 400, and 500, 600, 700 were used in this study. 
Remember that the goal is to find the smallest $n$ which is sufficient for proper coverage.

@@

Only $CI_{\hat{\theta}_{n}}$ managed to solve the deterioration problem with respect to recovering $\phi$, indicating that this interval 
should be used as opposed to $CI_{BCA, \hat{\theta}_{n}}$ when estimating $\phi$. As expected, a higher $\phi$ demands a 
larger $n$ to estimate all parameters, but a negative $\phi$ may only require a lesser $n$. For example, for both $\phi = 0.2$ and $0.4$, 
$n = 100$ was sufficient to recover $\mu$ at a rate of approximately 95\%.

@@

The motivation for the study is the idea that block bootstrap procedure will perfectly estimate a parameter of a time series given an infinitely large sample. 
The goal for this study was to find what is a good enough finite sample length for block bootstrap estimation to recover the parameter of a 
time series at an acceptable rate. Assuming that $CI_{\hat{\theta}_{n}}$ is used (as it was found to work better overall, especially when estimation $\phi$), 
the results of this study suggest that even for estimating $\mu$, an $n$ of around 500 may be necessary if $\phi$ is unknown. 
An $n$ of around 400 may be necessary to estimate $\sigma$ is $\phi$ is unknown. If $\phi$ is already known, 
a lesser $n$ may be adequate to estimate these parameters. When $\phi$ was negative, 
a $n$ of around 100 was sufficient to estimate $\mu$. However, in real world applications, $\phi$ is more commonly found to be positive, 
so a larger $n$ is more likely to be necessary. Lastly, to estimate $\phi$, an $n$ of around 400 may be necessary. 
This information could prove to be useful for researching using block bootstrap estimation of time series in domains such as econometrics. 
Future studies could investigate if there are types of block bootstrap intervals that can be constructed that 
would more appropriate recover the parameters of a time series. One could also investigate the $n$ needed to 
make inferences about other forms of serially dependent data such as an MA process.
